---
title: '#TidyTuesday'
author: "Amanda Luby"
date: "Week of 5/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "plots/")
library(tidyverse)
library(ggrepel)
library(grid)
library(ggtext)
```


```{r}
vb_matches <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-19/vb_matches.csv', guess_max = 76000)
```

First thing is to create a new dataset to find summaries for each pair.

```{r}
pairs_wins <- vb_matches %>%
  select(circuit, tournament, year, gender, match_num, w_player1, w_player2, l_player1, l_player2) %>%
  pivot_longer(., c(w_player1, l_player1), values_to = "player_name") %>%
  separate(name, c("outcome", "player_index")) %>%
  mutate(
    partner = ifelse(outcome == "w", w_player2, l_player2)
  ) %>%
  unite(., team, c(player_name, partner), remove = FALSE) %>%
  unite(., team_rev, c(partner, player_name)) %>%
  select(circuit, tournament, year, gender, match_num, outcome, team, team_rev) %>%
  #pivot_longer(., c(team, team_rev), values_to = "team") %>% # uncomment to double-add teams
  #select(-name) %>%
  group_by(team, year) %>%
  summarize(
    team_rev = team_rev[1],
    gender = gender[1],
    total_tourn = length(unique(tournament)),
    wins = sum(outcome == "w"),
    win_pct = mean(outcome == "w")
  ) %>%
  separate(team, c("player1", "partner"), sep = "\\_", remove = FALSE)
```

Created a list of olympic medalist teams (from Wikipedia). 

```{r}
olympic_medalists = c(
  "Ricardo Santos_Emanuel Rego",
  "Pablo Herrera_Javier Bosma",
  "Stefan Kobel_Patrick Heuscher",
  "Phil Dalhausser_Todd Rogers",
  "Marcio Araujo_Fabio Luiz Magalhaes",
  "Julius Brink_Jonas Reckermann",
  "Alison Cerutti_Emanuel Rego",
  "Martins Plavins_Janis Smedins",
  "Alison Cerutti_Bruno Oscar Schmidt",
  "Daniele Lupo_Paolo Nicolai",
  "Alexander Brouwer_Robert Meeuwsen",
  "Misty May-Treanor_Kerri Walsh Jennings",
  "Shelda Bede_Adriana Behar",
  "Holly McPeak_Elaine Youngs",
  "Jia Tian_Jie Wang",
  "Chen Xue_Xi Zhang",
  "Jennifer Kessy_April Ross",
  "Juliana Felisberta_Larissa Franca",
  "Laura Ludwig_Kira Walkenhorst",
  "Agatha Bednarczuk_Barbara Seixas",
  "April Ross_Kerri Walsh Jennings"
)
```

I'll restrict to pairs that have more than 10 wins, then add another variable to designate whether the team has medaled at the olympics. (Note that I check both orders of player names)
```{r}
pairs_wins %>%
  filter(wins >10) %>%
  mutate(
    medalist = ifelse(team %in% olympic_medalists, TRUE, ifelse(team_rev %in% olympic_medalists, TRUE, FALSE))
  ) -> pairs_data
```

Let's see whether we can see any trends by looking at team performance across years, split by M/F:

```{r win_pct_by_year}
pairs_data %>%
  #filter(medalist == TRUE) %>%
  ggplot(., aes(x = year, y = win_pct, size = wins, col = medalist, grp = team)) +
  geom_point(alpha = .4) +
  geom_line(size = .5, alpha = .5) + 
  scale_color_manual(values=c("#999999", "#3e0b80")) + 
  theme_minimal() +
  theme(legend.position = "bottom") +
  facet_grid(cols = vars(gender)) 

```
Medalists have higher win percentages, but it's hard to really see what's going on across time (eg do olympic medalist teams play together longer?)

Let's make another summary by pair, but now create a *single point* for each team, with the x-axis being the number of years they've played together, and the y-axis as their overall win percentage. 

```{r rough_win_pct_by_years_together}
pairs_data %>%
  ungroup() %>%
  group_by(team) %>%
  summarize(
    gender = gender[1],
    total_tourn = sum(total_tourn),
    total_wins = sum(wins),
    win_pct = sum(wins)/sum(wins/win_pct),
    years = length(unique(year)),
    medalist = medalist[1]
  ) -> pairs_data_summary

mean_winpct= pairs_data_summary %>%
  ungroup() %>%
  group_by(gender, medalist) %>%
  summarize(win_pct = mean(win_pct))

mean_years= pairs_data_summary %>%
  ungroup() %>%
  group_by(gender, medalist) %>%
  summarize(years = mean(years))

pairs_data_summary %>%
  ggplot(., aes(x = years, y = win_pct, size = total_wins, col = medalist, label = ifelse(medalist, team, ""))) + 
  geom_jitter(alpha = .65) +
  geom_hline(aes(yintercept = win_pct, col = medalist), mean_winpct, lty=2) + #add group averages for win_pct
  geom_vline(aes(xintercept = years, col = medalist), mean_years, lty=2) + #add group averages for years
  scale_color_manual(values=c("#999999", "#3e0b80")) + 
  theme_minimal() +
  theme(legend.position = "bottom") +
  facet_grid(cols = vars(gender))  + 
  geom_text_repel(size = 2) 
```

Much better! Now let's make it pretty: 


```{r final_win_pct_by_years_together}
pos <- position_jitter(width = 0.3, seed = 2)
pairs_data_summary %>%
  mutate(gender = ifelse(gender == "M", "Men", "Women"),
         team = gsub("_", " &\n", team)) %>%
  ggplot(., aes(x = years, y = win_pct, size = total_wins, col = medalist, label = ifelse(medalist, team, ""))) + 
  geom_point(position = pos, alpha = .6) +
  geom_point(data = pairs_data_summary[pairs_data_summary$medalist,], position = pos, col = "#3e0b80") %>%
  #geom_hline(aes(yintercept = win_pct, col = medalist), mean_winpct, lty=2) + #uncomment to add group averages for win_pct
  #geom_vline(aes(xintercept = years, col = medalist), mean_years, lty=2) + #uncomment to add group averages for years
  scale_color_manual(values=c("#999999", "#3e0b80")) + 
  theme_minimal(base_family = "Optima") +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_markdown(lineheight = 1.1),
        plot.subtitle = element_markdown(lineheight = 1.1),
        plot.caption = element_markdown(size = 11)) +
  facet_grid(cols = vars(gender))  + 
  geom_text_repel(position = pos,
                  size = 2, 
                  force = 8, 
                  point.padding = 0, 
                  box.padding = .5, 
                  family = "Optima", 
                  min.segment.length = 0,
                  xlim = c(-Inf, Inf), 
                  ylim = c(-Inf, Inf)) + 
  coord_cartesian(clip = "off") +
  labs(
    title = "**Beach Volleyball 2002-2019**",
    subtitle ="<span style='font-size:11pt'>Years Played and Win Percentage for **<span style='color:#3e0b80;'>Olympic Medalist</span>** and **<span style='color:#999999;'>Non-Medalist</span>** teams</span>",
    x = "Years Played Together", 
    y = "Win Percentage",
    caption = "#TidyTuesday 5/19/2020"
  ) +
  guides(color = FALSE)
```